---
title: "Lab 02: Data Structures in R"
author: "Bruno Grande and Santina Lin"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

## Tutorial

### Learning Objectives

- Describe the main vector types (logical, integer, double, character)
  - Define vectors of length 1 for each type (without the `c()` function)
  - Define longer vectors using the `c()` function
  - Create sequences of integers using the colon (`:`) notation
  - Explain how vector types are enforced (type coercion)
  - Determine the type of an existing vector
  - Explain vectorization in R and why it's powerful
  - Describe how R recycles values
  - Differentiate and convert between character vectors and factors
- Describe higher-level data structures (lists, matrices, data frames)
  - Provide example use cases for data frames versus matrices
  - Explain how data frames are related to lists
  - Examine a data frame using `str()` and `head()`
  - Differentiate between the output of `typeof()` and `class()`
- Subset vectors and data structures using the `[` operator
  - Explain the implications of 1-based indexing in R
  - Select elements/rows/columns by position or name
  - Remove elements/rows/columns by position
  - Subset elements/rows/columns using logical operations or vectors
  - Differentiate between the `[` and the `[[` subsetting operators

### Context

In the previous tutorial, you learned how to navigate the RStudio interface and ran your first bits of R code. You should now be familair with operators like `+` (mathematical) and `==` (logical) as well as functions like `sqrt()`. Don't worry: you don't have to remember every function and their respective arguments (_i.e._ inputs). You can use Google and the R help pages for this, like every other data scientist. 

Today, we're going to take a look at what data looks like in R. Each programming language organizes their data in different ways. These are known as data structures. This tutorial will focus on the most important data structure in R, the data frame, and its key component, the vector. These data structures are fundamental to data science in R, so pay close attention.

You will be exploring the Canadian COVID-19 statistics for the first six months of 2020. The data was originally downloaded from the [EU Open Data Portal](https://data.europa.eu/euodp/en/data/dataset/covid-19-coronavirus-data). It was lightly processed for this tutorial. 

### The Data Frame

#### Loading the Data Frame

Let's start by loading the data. You will learn more about this process in a future tutorial. For now, you are going to simply load the data from the Internet and preview it. The `read.csv()` function can take a file path or a URL and returns a data frame. Let's see it in action!

As you can see, a data frame is nothing more than a two-dimensional table consisting of rows and columns. This one has six rows (one for each month) and five columns. The columns contain different types of information: either integers or text.

In many respects, data frames look like Excel tables. However, data frames in R are strictly rectangular. Every column has the same length, and the same rule applies to rows. This doesn't mean you can't have missing data. Rather, you will see missing data indicated by the value `NA` (not seen here in `covid`).

```{r}
covid <- read.csv("https://bit.ly/cov-ca")

covid
```

The class is another way to refer to what kind of data structure is stored in a variable. Here, we can confirm that the `covid` variable contains a data frame.

```{r}
class(covid)
```

#### Exploring the Data Frame

As we saw in the previous tutorial, functions can be very useful. In fact, there are a number of functions that work well with data frames. Notably, the `nrow()` and `ncol()` functions can be used to count the number of rows and columns in the input, respectively. Admittedly, this is more useful with larger datasets.

```{r}
nrow(covid)

ncol(covid)
```

Another useful function is called `summary()` which will display statistics for numeric columns and tabulate values in text columns.

```{r}
summary(covid)
```

The `head()` and `tail()` functions are useful for extracting the first or last few rows in a data frame. These are especially useful for larger data frames.

```{r}
head(covid, 2)

tail(covid, 2)
```

### The Vector

#### Performing Operations on the Vector

That's enough boring stuff for now. Let's start actually handling the data. The most basic operation you can do with a data frame is extracting a column. There are several ways of doing this, but the simplest is the `$` operator. For example, if we want to extract the `cases` column, which contains the number of new COVID-19 cases during each month in Canada. You can clearly see the rapid increase in cases and eventual plateau once social distancing slowed down the outbreak.

```{r}
covid$cases
```

It might be surprising to see the column laid out horizontally. That's because it's no longer considered a column of the data frame after we use the `$` operator. Specifically, columns consist of a data structure known as a vector.

```{r}
typeof(covid$cases)
```

vectorization

```{r}
covid$cases / 1000
```

```{r}
covid$cases_1000 <- covid$cases / 1000

covid
```

```{r}
paste("Canada had", covid$cases_1000, "thousand cases in", covid$month)
```

```{r, error=TRUE}
covid$month + covid$cases_1000
```

#### Vector Types

type abbreviations

```{r}
str(covid)
```

```{r}
new_vector <- vector(mode = "character", length = 3)

new_vector
```

```{r}
typeof(new_vector)
```

```{r}
class(new_vector)
```

```{r}
str(new_vector)
```

```{r}
?vector
```

Concatenation

The `[1]` just indicates that the number appearing to its right is the first value of the output. As you will see later, the output of statements in R is not restricted to single values.

```{r}
lgl_vector <- c(TRUE, FALSE, FALSE)

lgl_vector

typeof(lgl_vector)
```

```{r}
int_vector <- c(1L, 5L, 8L)

int_vector

typeof(int_vector)
```

numeric and double

```{r}
num_vector <- c(1, 5.8, 101.01)

num_vector

typeof(num_vector)
```

```{r}
chr_vector <- c("foo", "bar")

chr_vector

typeof(chr_vector)
```

#### Type Coercion

Mixing types
little difference between integer and numeric/double
Integers can be specified with the `L` suffix, but this is rarely needed in practice

```{r}
mixed_vector <- c(FALSE, TRUE)
mixed_vector

mixed_vector <- c(mixed_vector, 2L)
mixed_vector

mixed_vector <- c(mixed_vector, 3.0)
mixed_vector

mixed_vector <- c(mixed_vector, "four")
mixed_vector
```

```{r}
as.numeric(mixed_vector)

as.logical(mixed_vector)

as.logical(as.numeric(mixed_vector))
```

```{r}
covid$cases_1000 <- as.integer(covid$cases_1000)

covid
```

### The Factor

```{r}
months <- factor(covid$month)

months
```

Hints at integer nature

```{r}
class(months)
str(months)
```

integer?

```{r}
typeof(months)
```

```{r}
as.character(months)

as.integer(months)

levels(months)
```

```{r}
factor(covid$month, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun"))
```

### The List

```{r}
list_example <- list(FALSE, TRUE, 2L, 3.0, "four")
list_example
```

```{r}
another_list <- list(title = "Numbers", numbers = 1:10, data = TRUE)
another_list
```

Same syntax as data frame. That's because data frames are where each column (vector or factor) has the same length

```{r}
another_list$numbers
```

```{r}
typeof(covid)
```

### The Matrix

```{r}
matrix_example <- matrix(1:24, ncol=6, nrow=4)
matrix_example
```

```{r}
class(matrix_example)
typeof(matrix_example)
```

Explain use cases for matrices vs data frames

### Subsetting

```{r}
x <- covid$cases
names(x) <- covid$month
x
```


#### Vector Subsetting Using Indices

```{r}
x[1]
x[4]
```

```{r}
x[c(1, 2, 3)]
x[1:3]
```

```{r}
x[c(1, 1, 3)]
```

```{r}
x[9]
```

```{r}
x[0]
```

```{r}
x[-2]
```

```{r}
x[c(-1, -5)]
x[-c(1, 5)]
```

Common mistake

```{r, error=TRUE}
x[-1:3]
```

```{r}
x[-(1:3)]
```

Remains unchanged so far

```{r}
x
```

```{r}
x <- x[-c(1, 6)]
x
```

#### Vector Subsetting Using Names

```{r}
x["Feb"]
x[c("Feb", "Apr")]
```

#### Vector Subsetting Using Logical Operations

Months with over 10000 cases

```{r}
x
x[c(FALSE, TRUE, TRUE, FALSE)]
```

```{r}
over10thousand <- x >= 10000
over10thousand
```

```{r}
x[over10thousand]
```

No need for intermediate variable

```{r}
x[x >= 10000]
```

AND

```{r}
x[x >= 30000 & x <= 40000]
```

OR

```{r}
x[x < 30000 | x > 40000]
```

==, not =

```{r}
names(x) == "May"
```

```{r}
x[names(x) == "May"]
```

Explain recycling

```{r}
x[names(x) == c("Feb", "Apr", "May")]
```

Look up help for `%in%`

```{r}
x[names(x) %in% c("Feb", "Apr", "May")]
```

`!` means NOT

```{r}
x[! names(x) %in% c("Mar")]
```

#### List Subsetting

```{r}
another_list
```

```{r}
another_list[2]

class(another_list[2])
```

```{r}
another_list[[2]]

class(another_list[[2]])
```

```{r}
another_list[1:2]
```

```{r, error=TRUE}
another_list[[1:2]]
```

```{r}
another_list[-(1:2)]
```

```{r, error=TRUE}
another_list[[-(1:2)]]
```

```{r}
another_list[c("numbers", "data")]
```

```{r}
another_list["numbers"]
```

```{r}
another_list[["numbers"]]
```

```{r}
another_list$numbers
```

#### Data Frame Subsetting

```{r}
covid
```

```{r}
covid$cases
covid[["cases"]]
covid[[4]]
```

```{r}
covid["cases"]
```

```{r}
covid[3, 4]
```

```{r}
covid[3,]
```

```{r}
covid[, 4]
```

## More Resources

- http://swcarpentry.github.io/r-novice-gapminder/04-data-structures-part1
- http://swcarpentry.github.io/r-novice-gapminder/06-data-subsetting
- http://swcarpentry.github.io/r-novice-gapminder/09-vectorization

## Assignment

No assignment for this lab.
