---
title: "Lab 07: Categorical Tests and Multiple Hypothesis Testing"
author: "Bruno Grande"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

## Tutorial

### Learning Objectives

- Perform independence tests on categorical data
  - Differentiate between Fisher's exact test and chi-squared test
  - Identify which test is most appropriate for a dataset
  - Visualize the categorical data for a dataset using the `ggplot` package
  - Process a data frame to produce a contingency table
  - Calculate P-value using `fisher.test()` and `chisq.test()`
- Perform P-value correction for multiple hypothesis testing
  - Visualize P-value distribution first using the `ggplot` package
  - Explain the multiple hypothesis problem
  - Correct P-values using `p.adjust()`
  - Detect anomalies in P-value distributions

### Context

### Exploring the data

```{r, warning=FALSE}
library(tidyverse)
library(broom)
library(patchwork)
```

```{r, warning=FALSE}
metadata <- read_csv("data/pannets_metadata.csv")

metadata
```

```{r}
rnaseq <- read_csv("data/pannets_expr_rnaseq.csv.gz")

rnaseq_long <- pivot_longer(rnaseq, cols = -Gene, 
                            names_to = "Tumour",
                            values_to = "Expr")

rnaseq_wide <- pivot_wider(rnaseq_long, id_cols = Tumour,
                           names_from = Gene, 
                           values_from = Expr)

head(rnaseq)
```

### Fisher's exact test

In the previous tutorial, we saw how we can compare continuous variables between two groups

In other words, one variable was continuous and the other was categorical (with only two groups)

In this tutorial, we will see how we can compare two categorical variables

Let's explore the patient metadata

```{r}
metadata
```

For small samples, we tend to use the Fisher's exact test

Before running the test, we need to consider the contingency table

Let's focus on the two columns for now: sex and metastasis

```{r}
select(metadata, Sex, Metastasis)
```

In this case, the contingency table counts up the number for each pair of values

In this case, there are two values for sex (male and female) and two values for metastasis (True or False)

Using the tidyverse, we could obtain those counts as follows, where `n()` counts the number of rows in each group

```{r}
metadata %>% 
  group_by(Sex, Metastasis) %>% 
  summarize(n = n())
```

The `count()` dplyr function is a shortcut for this common pattern

```{r}
count(metadata, Sex, Metastasis)
```

Normally, contingency tables are formatted in two dimensions, as follows

```{r}
count(metadata, Sex, Metastasis) %>% 
  pivot_wider(id_cols = Metastasis, values_from = n,
              names_from = Sex, names_prefix = "Sex_")
```

Visually, you can check if there is a trend for more metastases among males or females

The null hypothesis is that there is no difference between the proportion of males-to-females for cases with metastases and those without, or vice versa

We can calculate these proportions

Based on this, it seems that females are more likely to have metastases, but is this significant?

This is where the Fisher's exact test come in

```{r}
count(metadata, Sex, Metastasis) %>% 
  pivot_wider(id_cols = Metastasis, values_from = n,
              names_from = Sex, names_prefix = "Sex_") %>% 
  mutate(Proportion = Sex_F / Sex_M)
```

The `fisher.test()` does the counting for you if you provide two "parallel" vectors consisting of categorical values

Notice that the order doesn't matter

The P-value is 0.5, so the trend we noticed isn't significant, presumably due to the small sample size

```{r}
fisher.test(metadata$Sex, metadata$Metastasis)
fisher.test(metadata$Metastasis, metadata$Sex)
```

Let's look at another pair of categorical variables: subtype and metastasis

Here, we are using alternative approaches to create the contingency table

`table()` works like `fisher.test()` by taking in two "parallel vector"

```{r}
table(Subtype = metadata$Subtype, Metastasis = metadata$Metastasis)
```

On the other hand, `xtabs()` works off of a data frame, but has an unusual interface

The `~ Subtype + Metastasis` part is called a formula, which we will revisit later when we tackle linear regression

The main advantage of `xtabs()` is its conciseness

```{r}
xtabs(~ Subtype + Metastasis, metadata)
```

From the contingency tables above, you can see whether an association (positive or negative) exists between the variables if there is a relative abundance or depletion along the diagonal

Here, you can see that the diagonal has relatively few cases (4 + 4 = 8) compared to the rest of the table

Let's verify with a statistical test, providing the contingency table directly to `fisher.test()`

In this case, the association is significant (P-value = 0.006)

In other words, based on the contingency table, we can say that the A-D-M mutant cases are significantly more likely to develop metastases

Importantly, we cannot say that the mutations cause metastases: correlation does not imply causation

But this generates a hypothesis that can be investigated with additional experiments

```{r}
fisher.test(xtabs(~ Subtype + Metastasis, metadata))
```

Let's look how we can visualize these results

There are only a few ways of visualizing contingency tables

First, you can encode the counts as the point size

```{r}
ggplot(metadata, aes(x = Subtype, y = Metastasis)) +
  geom_count()
```

Second, there is the most customized mosaic plot

For this, we need to install a ggplot2 extension package

```{r}
library(ggmosaic)

ggplot(metadata) +
  geom_mosaic(aes(x = product(Metastasis, Subtype), fill = Metastasis))
```

A mosaic plot looks like a stacked relative barplot (adding up to 100%), but the width of the bars is proportional to the sample size of that bar

For example, from the plot below, we can see that there are more A-D-M mutant cases than WT cases (see numbers below)

From this plot, we can see how the "metastasis rate" is much higher in mutant cases

```{r}
table(metadata$Subtype)
```

However, it might make more sense to put the metastasized cases at the bottom and set the WT group as the left-most reference group

```{r}
metadata %>% 
  mutate(Metastasis = as.character(Metastasis), 
         Metastasis = fct_relevel(Metastasis, "TRUE", "FALSE"),
         Subtype = fct_relevel(Subtype, "A-D-M WT")) %>% 
  ggplot() +
  geom_mosaic(aes(x = product(Metastasis, Subtype), fill = Metastasis)) +
  scale_fill_discrete(limits = c("FALSE", "TRUE"))
```

### Chi-squared test

###  Multiple hpothesis testing

## More Resources

- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5426219
- http://varianceexplained.org/statistics/interpreting-pvalue-histogram/

## Assignment
